Improved Approximation Algorithms for the Maximum Carpool Matching Problme

### 2017-02-15 src/local.tex
merge dror

In this section we present a $(\half + \inv{2c_{\max}}
- \eps)$-approximation algoithm for unweighted \carpool whose running
time is polynomial if $c_{\max} = O(1)$.

Let $k$ be an integer to be determined later.  The local search
algorithm maintains a feasible matching throughout its execution and
operates in iterative manner where in each iteration it tries to find
a better solution by replacing a subset of at most $k$ edges in the
current solution with another (larger) subset of edges not in the
solution.  The local search algorithm halts when no improvement can be
done.  The algorithm is described in Algorithm~\ref{alg:local}.

\begin{algorithm}
\SetKw{True}{true}
\SetKw{False}{false}
\KwIn{$G = (V, A)$, $c : V \rightarrow \N$, $k$}
\KwOut{$M$}
$M \leftarrow \emptyset$								\\
\Repeat{done}{
\label{line:outerloop}
 	$done \leftarrow{}$ \True							\\
 	\ForAll{$M' \subseteq M : |M'| \leq k$}{
 		\ForAll{$A' \subseteq A \setminus M : |A'| = |M'| + 1$}{
			\If{$M \setminus M' \cup A'$ is feasible}{
				$M \leftarrow{} M \setminus M' \cup A'$	\\
				$done \leftarrow{}$ \False				\\
			}
		}
 	}
}
\Return{$M$}

\caption{Local Search}
\label{alg:local}
\end{algorithm}

We claim that the local search algorithm halts in polynomial time.
Observed that in every iteration the algorithm either improves the
value of the solution by one or otherwise it terminates.  Thus, after
a maximum of $n$ iterations the algorithm stops.  In every iteration
we examine all the subsets of edges of a fixed size and test for
feasibility, both these operations can be done in polynomial time.

@@

To analyze the approximation ratio achieved by the local search
algorithm consider an arbitrary but fixed optimal matching $M^*$.
Given the solution outputted by the local search algorithm, $M$, we
build the \emph{star graph} of the two solutions in which each node
represents a star from the optimal solution, namely from $\calS(M^*)$,
and an edge exists between two nodes if there is a star in $\calS(M)$
that intersects the two corresponding stars of the optimal solution.
%
Formally $H = (\calS(M^*), E)$ where
$
{E = \set{(S^*_i, S^*_j) : \exists S \in \calS(M), 
         V(S) \cap V(S^*_i) \neq \emptyset \land
         V(S) \cap V(S^*_j) \neq \emptyset}}
$.
Figure~\ref{fig:conflict} depicts a star graph.

@@

\begin{lemma}
The maximum degree of $H$ is $\cmax(\cmax+1)$.
\end{lemma}
\begin{proof}
Each star in $\calS(M^*)$ consists of at most $\cmax+1$ vertices and
each such vertex can belong to a star in $\calS(M)$ containing
additional $\cmax$ vertices, each of which is located in a different
star in $\calS(M^*)$.
\end{proof}

Define $\deg_M(v) \eqdf \din[M](v) + \dout[M](v)$.  For a subset
$U \subseteq V$ of vertices define $\deg_M(U) \eqdf \sum_{v \in
U} \deg_M(u)$.

@@

Since $\calT$ induces a connected subgraph of $H$, the removal of
edges in $M$ that intersect $V(\calT)$ decreased $\abs{M}$ by at most
$\deg_M(V(\calT)) - \abs{\calT} + 1$.
%
On the other hand, the increase in size is exactly
 $\half \deg_{M^*}(V(\calT)) \leq \cmax \abs{\calT}$.
%
Since $\abs{A(\calT)} \leq k$, we know that this difference can not be
positive, or else, the local search algorithm would not have been
terminated.  Thus 
\(
\half \deg_{M^*}(V(\calT)) \leq \deg_M(V(\calT)) - \abs{\calT} + 1
\),
and so
\[
\frac{\deg_M(V(\calT))}{\deg_{M^*}(V(\calT))}
\geq \half + \frac{\abs{\calT} - 1}{\deg_{M^*}(V(\calT))}
\geq \half + \frac{\abs{\calT} - 1}{2\cmax \abs{\calT}}
=    \half + \inv{2\cmax} - \inv{2\cmax \abs{\calT}}
~,
\]
as required.
\end{proof}

\begin{lemma}
\label{lemma:dec}
Let $G = (V,E)$ be an undirected connected graph with maximum degree
$\Delta$, and let $\ell < \abs{V}$.  
Then $G$ can be decomposed into connected components of size at least $\ell$,
and at most $\Delta \ell$.
\end{lemma}

\begin{proof}
We give a constructive proof.
If $\ell \leq |V| \leq \Delta \ell$ we are done.
%
Otherwise, remove (in any desired way) up to $\ell$ connected vertices $C =
\{v_1,\ldots,v_m\}$ from $G$ until $G$ is disconnected.
If $m = \ell$ then let $C_1 = \{v_1,\ldots,v_\ell\}$ and recursively decompose
$G - C$.
If, on the other hand, the above process disconnects $G$ then consider the
connected components $\{C_1, \ldots, C_t\}$ of $G - C$.
Join all the small components and $C$, i.e. $C \cup \bigcup_{i : |C_i| < \ell}$
and decompose the other components recursively, 
noting that $|C \cup \bigcup_{i : |C_i| < \ell}| \leq \Delta \ell$.
If all the components are large, i.e. $\forall i |C_i| \geq \ell$ then join $C$
to an arbitrary component and decompose the resulting components recursively.

\iffalse
Let $T$ be any spanning tree of $G$ and let $r$ be an arbitrary root.
Starting with $r$, repeatedly choose a child whose subtree is strictly
larger than $\frac{\Delta-1}{\Delta} \abs{V}$, until this is not
possible.  Hence, we have reached a vertex $v$ whose subtree contains
more than $\frac{\Delta-1}{\Delta} \abs{V}$ vertices, but the subtrees
of its children contain at most $\frac{\Delta-1}{\Delta} \abs{V}$
vertices.
%
If $v = r$, then $v$ has a child $u$ with at least $\inv{\Delta}
(\abs{V}-1) \geq \ell$ vertices.  Otherwise, there exists a child $u$
of $v$ whose subtree is of size at least
$\inv{\Delta} \abs{V} \geq \ell$.  Disconnect $u$ and the vertices in
its subtree from the graph.
%
In both cases $u$'s subtree contains at most
$\frac{\Delta-1}{\Delta} \abs{V}$ vertices, and thus at least
$\inv{\Delta} \abs{V} \geq \ell$ vertices remain in the graph.  Hence
we obtain two connected subgraphs of size at least $\ell$.
%
We repeat this procedure recursively on any component of size
strictly larger $\Delta\ell$.
\fi
\end{proof}

@@

@@

It remains to consider a maximal component $\calT$ such that If
$\abs{M \cap A(\calT)} > k$.  Observe that since the number of edges
in $S \in \calS(M^*)$ is at most $\cmax$, it must be that
$\abs{V(\calT)} > \frac{k}{\cmax}$.
%
Due to Lemma~\ref{lemma:dec} (with $\ell = \frac{k}{\cmax}$) we can
partition $\calT$ into connected components each of which contains
between $\frac{k}{\cmax^2(\cmax+1)}$ and $\frac{k}{\cmax}$ vertices.
%
Since each such vertex set $\calX$ is connected and contains at most
$\frac{k}{\cmax}$ stars, it follows that $\abs{A(\calX)} \leq k$.
%
Due to Lemma~\ref{lemma:r} we have that
\[
\frac{\deg_M(V(\calX))}{\deg_{M^*}(V(\calX))} 
\geq \half + \inv{2\cmax} - \frac{\cmax(\cmax+1)}{2k}
~.
\]
Since $\deg_{M^*}(V(\calT)) = \sum_{\calX} \deg_{M^*}(V(\calX))$ and
$\deg_M(V(\calT)) = \sum_{\calX} \deg_M(V(\calX))$, it follows that
\[
\frac{\deg_M(V(\calT))}{\deg_{M^*}(V(\calT))}
\geq \half + \inv{2\cmax} - \frac{\cmax(\cmax+1)}{2k}
~,
\]
and thus
\[
\abs{M \cap A(\calT)}
\geq \paren{ \half + \inv{2\cmax} -
             \frac{\cmax(\cmax+1)}{2k}} \abs{M^* \cap A(\calT)}
~,
\]
and the lemma follows.
\end{proof}

@@

\begin{corollary}
\label{cor:local}
There exists a $(\half + \inv{2\cmax} - \eps)$-approximation algorithm
for \carpool, for every $\eps>0$.
\end{corollary}

In the appendix we show that our analysis is tight.

>>>>

In this section we present a local search $(\half + \inv{2c_{\max}}
- \eps)$-approximation algoithm for unweighted \carpool whose running
time is polynomial if $c_{\max} = O(1)$.

Let $k$ be a constant integer to be determined later.
Algorithm~\textbf{EdgeSwap} (Algorithm~\ref{alg:local}) maintains a
feasible matching $M$ throughout its execution and operates in
iterative manner where in each iteration it tries to find a better
solution by replacing a subset of at most $k$ edges in the current
solution with another (larger) subset of edges not in the solution.
The algorithm halts when no improvement can be done.

\begin{algorithm}
\caption{\textbf{EdgeSwap}$(G,c,k)$}
\label{alg:local}
\begin{small}
\SetKw{True}{true}
\SetKw{False}{false}
%\KwIn{$G = (V, A)$, $c : V \rightarrow \N$, $k$}
%\KwOut{$M$}
$M \leftarrow \emptyset$								\\
\Repeat{done}{
\label{line:outerloop}
 	$done \leftarrow{}$ \True							\\
 	\ForAll{$M' \subseteq M : |M'| \leq k$}{
 		\ForAll{$A' \subseteq A \setminus M : |A'| = |M'| + 1$}{
			\If{$M \setminus M' \cup A'$ is feasible}{
				$M \leftarrow{} M \setminus M' \cup A'$	\\
				$done \leftarrow{}$ \False				\\
			}
		}
 	}
}
\Return{$M$}
\end{small}
\end{algorithm}

Algorithm~\textbf{EdgeSwap} terminates in polynomial time, since in
every non-final iteration it improves the value of the solution by
one.  Thus, after at most $n$ iterations the algorithm terminates.  In
every iteration the algorithm examines all subsets of edges of a fixed
size and tests for feasibility, both these operations can be done in
polynomial time.

@@

It remains to analyze the approximation ratio of \textbf{EdgeSwap}.
Consider an optimal matching $M^*$, and $M$ be the matching computed
by \textbf{EdgeSwap}.  Given both matchings we build the \emph{star
graph} of the two solutions in which each node represents a star from
the optimal solution, namely from $\calS(M^*)$, and an edge exists
between two nodes if there is a star in $\calS(M)$ that intersects the
two corresponding stars of the optimal solution.
%
Formally $H = (\calS(M^*), E)$ where
$
{E = \set{(S^*_i, S^*_j) : \exists S \in \calS(M), 
         V(S) \cap V(S^*_i) \neq \emptyset \land
         V(S) \cap V(S^*_j) \neq \emptyset}}
$.
Figure~\ref{fig:conflict} depicts a star graph.

@@

\begin{lemma}
The maximum degree of $H$ is $\cmax(\cmax+1)$.
\end{lemma}
\begin{proof}
Each star in $\calS(M^*)$ contains at most $\cmax+1$ vertices and each
such vertex can belong to a star in $\calS(M)$ containing additional
$\cmax$ vertices, each of which is located in a different star in
$\calS(M^*)$.
\end{proof}

In what follows we compare $\abs{M}$ and $\abs{M^*}$ in maximal
connected components of the star graph $H$.  Intuitively, we show that
$M$ is optimal on small maximal components, and that the approximation
ratio on medium (non-necessarily) components can be bounded due to the
termination condition of \textbf{EdgeSwap}.  Large maximal components
will be partitioned into medium components.

We first show that large connected graphs (or maximal connected
components) can be partitioned in to medium size components.  The
proof is given in Appnedix~\ref{sec:omitted}.

\begin{lemma}
\label{lemma:dec}
An undirected connected graph $G = (V,E)$ with maximum degree
$\Delta$, can be decomposed into connected components of size at least
$\ell$ and at most $\Delta \ell$, if $\ell \leq \abs{V}$.
\end{lemma}

\iffalse %%%%%% Removed

\begin{proof}
Let $T$ be any spanning tree of $G$ and let $r$ be an arbitrary root.
Starting with $r$, repeatedly choose a child whose subtree is strictly
larger than $\frac{\Delta-1}{\Delta} \abs{V}$, until this is not
possible.  Hence, we have reached a vertex $v$ whose subtree contains
more than $\frac{\Delta-1}{\Delta} \abs{V}$ vertices, but the subtrees
of its children contain at most $\frac{\Delta-1}{\Delta} \abs{V}$
vertices.
%
If $v = r$, then $v$ has a child $u$ with at least $\inv{\Delta}
(\abs{V}-1) \geq \ell$ vertices.  Otherwise, there exists a child $u$
of $v$ whose subtree is of size at least
$\inv{\Delta} \abs{V} \geq \ell$.  Disconnect $u$ and the vertices in
its subtree from the graph.
%
In both cases $u$'s subtree contains at most
$\frac{\Delta-1}{\Delta} \abs{V}$ vertices, and thus at least
$\inv{\Delta} \abs{V} \geq \ell$ vertices remain in the graph.  Hence
we obtain two connected subgraphs of size at least $\ell$.
%
We repeat this procedure recursively on any component of size
strictly larger $\Delta\ell$.
\end{proof}

\fi %%%%%% End removed

Define $\deg_M(v) \eqdf \din[M](v) + \dout[M](v)$.  For a subset
$U \subseteq V$ of vertices define $\deg_M(U) \eqdf \sum_{v \in
U} \deg_M(u)$.
%
Observe that $\abs{M} = \half \deg_M(V)$.

In the next lemma we bound the degree ratio in a component that
contains stars with at most $k$ arcs.

@@

Since $\calT$ induces a connected subgraph of $H$, the removal of
edges in $M$ that intersect $V(\calT)$ decreased $\abs{M}$ by at most
$\deg_M(V(\calT)) - \abs{\calT} + 1$.
%
On the other hand, the increase in size is exactly
 $\half \deg_{M^*}(V(\calT)) \leq \cmax \abs{\calT}$.
%
Since $\abs{A(\calT)} \leq k$, we know that this difference can not be
positive, or else, \textsc{EdgeSwap} would not have terminated.  Thus
\(
\half \deg_{M^*}(V(\calT)) \leq \deg_M(V(\calT)) - \abs{\calT} + 1
\),
and so
\[
\frac{\deg_M(V(\calT))}{\deg_{M^*}(V(\calT))}
\geq \half + \frac{\abs{\calT} - 1}{\deg_{M^*}(V(\calT))}
\geq \half + \frac{\abs{\calT} - 1}{2\cmax \abs{\calT}}
=    \half + \inv{2\cmax} - \inv{2\cmax \abs{\calT}}
~,
\]
as required.
\end{proof}

@@

It remains to bound the approximation ratio of \textbf{EdgeSwap}.

@@

It remains to consider a maximal component $\calT$ such that If
$\abs{M \cap A(\calT)} > k$.  Since the number of edges in
$S \in \calS(M^*)$ is at most $\cmax$, it must be that $\abs{V(\calT)}
> \frac{k}{\cmax}$.
%
Due to Lemma~\ref{lemma:dec} (with $\ell
= \frac{k}{\cmax^2(\cmax+1)}$) we can partition $\calT$ into connected
components each of which contains between $\frac{k}{\cmax^2(\cmax+1)}$
and $\frac{k}{\cmax}$ vertices.
%
Since each such vertex set $\calX$ is connected and contains at most
$\frac{k}{\cmax}$ stars, it follows that $\abs{A(\calX)} \leq k$.
%
Due to Lemma~\ref{lemma:r} we have that
\[
\frac{\deg_M(V(\calX))}{\deg_{M^*}(V(\calX))} 
\geq \half + \inv{2\cmax} - \frac{\cmax(\cmax+1)}{2k}
~.
\]
Since $\deg_{M^*}(V(\calT)) = \sum_{\calX} \deg_{M^*}(V(\calX))$ and
$\deg_M(V(\calT)) = \sum_{\calX} \deg_M(V(\calX))$, it follows that
\[
\frac{\deg_M(V(\calT))}{\deg_{M^*}(V(\calT))}
\geq \half + \inv{2\cmax} - \frac{\cmax(\cmax+1)}{2k}
~,
\]
and thus
\(
\abs{M \cap A(\calT)}
\geq \paren{ \half + \inv{2\cmax} -
             \frac{\cmax(\cmax+1)}{2k}} \abs{M^* \cap A(\calT)}
%~,
\).
%and the lemma follows.
\end{proof}

@@

\begin{corollary}
\label{cor:local}
There exists a $(\half + \inv{2\cmax} - \eps)$-approximation algorithm
for unweighted \carpool, for every $\eps>0$.
\end{corollary}

In the Appendix~\ref{sec:tight} we show that our analysis is tight.
